Blue Hunters: Bluetooth RSSI Locator Robots

### 2018-02-02 article/article.tex
start rewriting

            pdftitle={Blue Hunters: Bluetooth RSSI Locator Robots},
            pdfauthor={Jacob Glueck (jng55); Jane Du (zd53); Justin Cray (jgc232)},
            pdfborder={0 0 0},
            breaklinks=true}

@@

@@

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{Blue Hunters: Bluetooth RSSI Locator Robots}
\providecommand{\subtitle}[1]{}
\subtitle{\emph{ECE 4760, Fall 2017}}
\author{Jacob Glueck (\href{mailto:jng55@cornell.edu}{jng55}) \and Jane Du (\href{mailto:zd53@cornell.edu}{zd53}) \and Justin Cray (\href{mailto:jgc232@cornell.edu}{jgc232})}
\date{December 6, 2017}

@@

Position estimation and navigation for robots is an interesting problem
in the field of robotics. We investigate the challenge of determining
relative position using signal strength; by placing a moving robot on
the same plane as a beacon, oriented and positioned randomly, the
robot can hunt out and move towards the beacon by itself. We discuss
our procedures and challenges we encountered in this article.

Specifically, we built 2 small cars which used Bluetooth Received Signal Strength
Indicator (RSSI) measurements to navigate towards a stationary base
station. The cars and base station used a Bluetooth Low Energy (BLE) 4.0
module to take the measurements and a PIC32MX250 microcontroller. The
cars also used a 3 axis magnetometer as as compass in order to reliably
turn, as well as 2 micro 9 g servos to drive. Each unit was powered with
3 AA batteries. Finally, the chassis and wheels of each car were 3D
printed.

@@

The design of the project can be split into 3 main parts: the chassis, the electronics (hardware), and the software. The following sections discuss implementation details.

@@

The robots are made from 4 3D printed pieces: 2 wheels, the frame, and
the caster in the back. The servos, a 3 AA battery holder, and a
perfboard containing all the circuitry are mounted directly to the
frame.

The robots where designed in \href{http://www.openscad.org/}{OpenSCAD},
and their source code is available in
\href{https://github.com/orangeturtle739/bluehunters/tree/master/cad}{our
git repository} and
\protect\hyperlink{appendix-b-source-listing}{below}. There are three
files, \href{generated/frame.scad.html}{\texttt{frame.scad}},
\href{generated/drag.scad.html}{\texttt{drag.scad}}, and
\href{generated/wheel.scad.html}{\texttt{wheel.scad}}, for each of the
three parts. The main modules are defined in
\href{generated/main.scad.html}{\texttt{main.scad}}, and the other files
just instantiate them. The following renderings show each part:

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{frame.png}
\caption{Robot chassis}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{drag.png}
\caption{Caster wheel on back of robot}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{wheel.png}
\caption{Robot wheel}
\end{figure}

The parts were printed in ABS using
\href{https://www.monoprice.com/product?c_id=107\&cp_id=10724\&cs_id=1072403\&p_id=13860\&seq=1\&format=2}{Maker
Select 3D Printer v2} printers. All parts were printed with a layer
height of 0.3 mm, as there was no need for a smooth finish or high
tolerances. The parts where sliced with
\href{https://ultimaker.com/en/products/ultimaker-cura-software}{Cura}.

\hypertarget{pictures}{%
\subsubsection{Pictures}\label{pictures}}

Full system (base station and 2 robots):

\includegraphics[width=0.5\textwidth,height=\textheight]{full_system.jpg}

Multiple views of the robots:

\begin{longtable}[]{@{}ll@{}}
\toprule
\includegraphics{top_view.jpg} &
\includegraphics{side_view.jpg}\tabularnewline
\includegraphics{front_view.jpg} &
\includegraphics{inside_view.jpg}\tabularnewline
\bottomrule
\end{longtable}

@@

\hypertarget{bluetooth-modules}{%
\paragraph{Bluetooth modules}\label{bluetooth-modules}}

The HM-10 bluetooth modules we bought off Ebay were fakes: they were not
made by Jnhuamao \footnote{Jnhuamao is the company which makes the HM-10
  module, along with other Bluetooth modules. Their English website is:
  \url{http://www.jnhuamao.cn/bluetooth.asp?id=1}.}, and did not come
with genuine Jnhuamao firmware. However, they did have real TI CC2541
chips. We realized they were fakes when they did not behave according to
the Jnhuamao data sheet. \footnote{The original data sheet comes from \cite{jnhuamaodatasheet}. A PDF is available at \cite{jnhuamaomit}, which may be out of date, but is eaiser to get.}
Luckily, the hardware on the fake chips is the same as
that of the genuine chips, minus an external crystal, and the genuine
firmware checks for the presence of the crystal, and works even without
it. \cite{crystal}
As such, we reprogrammed the chips with the genuine firmware according
to an
\href{http://forum.arduino.cc/index.php?topic=393655.msg2709528\#msg2709528}{Arduino
forum post}  \cite{crystal}:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We soldered wires to the programming pins on the breakout boards, and
  connected those pins to an
  \href{https://www.pjrc.com/teensy/teensy31.html}{Arduino Teensy 3.2}.
  We chose a Teensy because it is 3.3 V as opposed to 5, which would
  damage the 3.3 V CC2541.

  The pins were connected as follows:

@@

  The layout of the HM-10 board is:

  \includegraphics{hm10_pins.png}
\item
  We uploaded the
  \href{https://github.com/RedBearLab/CCLoader/blob/master/Arduino/CCLoader/CCLoader.ino}{\texttt{CCLoader.ino}
  sketch} to the Arduino.
\item
  Finally, we ran (in a Windows virtual machine)
  \href{https://github.com/RedBearLab/CCLoader/tree/master/Windows}{\texttt{CCLoader.exe}}.
  This program takes 3 arguments:

\begin{Shaded}
\begin{Highlighting}[]
\ExtensionTok{CCLoader.exe} \OperatorTok{<}\NormalTok{COM Port}\OperatorTok{>} \OperatorTok{<}\NormalTok{Firmware.bin}\OperatorTok{>}\NormalTok{ 0}
\end{Highlighting}
\end{Shaded}

  The firmware file came from the same Arduino form post, and can be
  found
  \href{http://forum.arduino.cc/index.php?action=dlattach;topic=393655.0;attach=183702}{here}.
\end{enumerate}

There is also an excellent
\href{https://www.youtube.com/watch?v=ez3491-v8Og}{YouTube video}, which
explains the firmware flashing process.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Connect the HM-10 module to a computer using a 3.3 V FTDI to USB
  adapter. Then, use PUTTY to establish a serial session (9600 baud,
  8N1). Send the chip \texttt{AT}; if it is connected properly, it will
  respond with \texttt{OK}.
\item
  Send the chip \texttt{AT+SBLUP} to put it in firmware update mode. It
  will respond with \texttt{OK+SBLUP}. Terminate the PUTTY session.
\item
  Run the \texttt{HMSoft.exe} program distributed in the firmware update
  download. We ran it in a Windows virtual machine because we did not
  trust it.
\item
  Select the proper port and firmware file using the software, and hit
  ``Load Image''. The software should handle the rest!
\item
  To make sure it worked, establish a serial connection again using
  PUTTY. Send \texttt{AT+VERS?} to query the chip for version
  information.
\end{enumerate}

\hypertarget{software}{%
\subsubsection{Software}\label{software}}

@@

\hypertarget{ble}{%
\paragraph{BLE}\label{ble}}

@@

\hypertarget{imu}{%
\paragraph{IMU}\label{imu}}

@@

\hypertarget{servos}{%
\paragraph{Servos}\label{servos}}

@@

\hypertarget{protothreads}{%
\paragraph{Protothreads}\label{protothreads}}

@@

\hypertarget{gradient-descent}{%
\paragraph{Gradient descent}\label{gradient-descent}}

@@

\hypertarget{testing}{%
\subsubsection{Testing}\label{testing}}

\hypertarget{ble-1}{%
\paragraph{BLE}\label{ble-1}}

@@

\hypertarget{imu-1}{%
\paragraph{IMU}\label{imu-1}}

@@

\hypertarget{servos-1}{%
\paragraph{Servos}\label{servos-1}}

@@

\hypertarget{whole-system}{%
\paragraph{Whole system}\label{whole-system}}

@@

\hypertarget{results}{%
\subsection{Results}\label{results}}

@@

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

@@

\hypertarget{appendices}{%
\subsection{Appendices}\label{appendices}}

\hypertarget{appendix-c-schematics}{%
\subsubsection{Appendix A: Data}\label{appendix-c-schematics}}

@@

\hypertarget{appendix-c-schematics}{%
\subsubsection{Appendix B: Schematics}\label{appendix-c-schematics}}

\includegraphics{schematic.png}

\hypertarget{appendix-d-bill-of-materials}{%
\subsubsection{Appendix D: Bill Of
Materials}\label{appendix-d-bill-of-materials}}

>>>>

    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=cyan,

@@

\usepackage{subfig}
\usepackage{float}

@@

\title{Blue Hunters: Bluetooth RSSI Locator Robots}
\author{Jacob Glueck (\href{mailto:jng55@cornell.edu}{jng55}) \and Jane Du (\href{mailto:zd53@cornell.edu}{zd53}) \and Justin Cray (\href{mailto:jgc232@cornell.edu}{jgc232})}
\date{December 6, 2017}

@@

Position estimation and navigation for indoor robots is an interesting problem.
In this article, we investigate using the Received Signal Strength Indicator (RSSI) of Bluetooth Low Energy (BLE) 4.0 chips to allow wheeled mobile robots to navigate towards a stationary base station.
Each robot was controlled by a PIC32MX250 microcontroller, and used a 3 axis magnetometer as as compass in order to reliably turn, as well as 2 micro 9 g servos to drive.
Each unit was powered with 3 AA batteries.
Finally, the chassis and wheels of each car were 3D printed.
Figure \ref{fig:robotsystem} shows the entire system.

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{full_system.jpg}
  \caption{Full system with 2 robots and the base station}
  \label{fig:robotsystem}
\end{figure}

@@

@@

The robots are made from 4 3D printed pieces: 2 wheels, the frame, and
the caster in the back.
Figure \ref{fig:robotchassis} shows the main components.
The servos, a 3 AA battery holder, and a perfboard containing all the circuitry are mounted directly to the frame.

\begin{figure}
  \centering
  \subfloat[Frame] {
    \includegraphics[width=0.5\textwidth,height=\textheight]{frame.png}
    \label{fig:robotfrane}
  }
  \subfloat[Wheel]{
    \includegraphics[width=0.5\textwidth,height=\textheight]{wheel.png}
    \label{fig:robotwheel}
  }
  \\
  \subfloat[Caster]{
    \includegraphics[width=0.5\textwidth,height=\textheight]{drag.png}
    \label{fig:robotcaster}
  }
  \caption{Components of robot chassis}
  \label{fig:robotchassis}
\end{figure}

The parts were printed in ABS using Maker Select 3D Printer v2 printers.
All parts were printed with a layer height of 0.3 mm, as there was no need for a smooth finish or high tolerances.
The parts where sliced with Cura.

\begin{figure}
  \centering
  \subfloat[Top view]{
    \includegraphics[width=0.5\textwidth]{top_view.jpg}
    \label{fig:robottop}
  }
  \subfloat[Side view]{
    \includegraphics[width=0.5\textwidth]{side_view.jpg}
    \label{fig:robotside}
  }
  \\
  \subfloat[Front view]{
    \includegraphics[width=0.5\textwidth]{front_view.jpg}
    \label{fig:robotfront}
  }
  \subfloat[Back view]{
    \includegraphics[width=0.5\textwidth]{inside_view.jpg}
    \label{fig:robotback}
  }
  \caption{Multiple views of the completed robots}
  \label{fig:robotviews}
\end{figure}

@@

Each robot consisted of 4 main components:
\begin{enumerate}
  \item PIC32MC250F128B microcontroller.
  \item MPU-9250 IMU with a 3 axis gryoscope, accelerometer, and magnetometer.
  \item HM-10 BLE module with a Texas Instruments CC2541 chip.
  \item Two 9 gram FS90R micro continuous rotation servos.
\end{enumerate}

The components are connected as shown in figure \ref{fig:schematic}.

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{schematic.png}
  \caption{Schematic}
  \label{fig:schematic}
\end{figure}

\paragraph{Bluetooth modules}

The HM-10 bluetooth modules we bought off Ebay were fakes: they were not made by Jnhuamao,
\footnote{Jnhuamao is the company which makes the HM-10 module, along with other Bluetooth modules. Their English website is: \url{http://www.jnhuamao.cn/bluetooth.asp?id=1}.}
and did not come with genuine Jnhuamao firmware.
However, they did have genuine TI CC2541 chips.
We realized they were fakes when they did not behave according to
the Jnhuamao data sheet.
\footnote{The original data sheet comes from \cite{jnhuamaodatasheet}.
A PDF is available at \cite{jnhuamaomit}, which may be out of date, but is eaiser to get.}
Luckily, the hardware on the fake chips is the same as that of the genuine chips, minus an external crystal, and the genuine firmware checks for the presence of the crystal, and works even without
it. \cite{crystal}
As such, we reprogrammed the chips with the genuine firmware according
to an Arduino \cite{crystal}:

\begin{enumerate}
\item
  We soldered wires to the programming pins on the breakout boards, and connected those pins to an Arduino Teensy 3.2.
  We chose a Teensy because it is 3.3 volts as opposed to 5, which would damage the 3.3 volt CC2541. The pins were connected as follows:

@@

  Figure \ref{fig:hm10} shows the layout of the HM-10 module.
  \begin{figure}
    \centering
    \includegraphics[width=0.3\textwidth]{hm10_pins.png}
    \caption{HM-10 module pin layout \cite{crystal}}
    \label{fig:hm10}
  \end{figure}

\item
  We uploaded the \texttt{CCLoader.ino} \cite{ccloader} sketch to the Arduino.
\item
  Finally, we ran (in a Windows virtual machine, due to the dubious origin of the software) \texttt{CCLoader.exe} \cite{ccloaderexe}.
  This program takes 3 arguments:

  \begin{Verbatim}[gobble=4]
    CCLoader.exe <COM Port> <Firmware.bin> 0
  \end{Verbatim}

  The firmware file came from the Arduino forum post \cite{crystal}, and can be
  found at \url{http://forum.arduino.cc/index.php?action=dlattach;topic=393655.0;attach=183702}.
\end{enumerate}

@@

\begin{enumerate}
\item
  Connect the HM-10 module to a computer using a 3.3 V FTDI to USB
  adapter. Then, use PUTTY to establish a serial session (9600 baud,
  8N1). Send the chip \texttt{AT}; if it is connected properly, it will
  respond with \texttt{OK}.
\item
  Send the chip \texttt{AT+SBLUP} to put it in firmware update mode. It
  will respond with \texttt{OK+SBLUP}. Terminate the PUTTY session.
\item
  Run the \texttt{HMSoft.exe} program distributed in the firmware update
  download. We ran it in a Windows virtual machine because we did not
  trust it.
\item
  Select the proper port and firmware file using the software, and hit
  ``Load Image''. The software should handle the rest!
\item
  To make sure it worked, establish a serial connection again using
  PUTTY. Send \texttt{AT+VERS?} to query the chip for version
  information.
\end{enumerate}

\subsubsection{Software}

@@

\paragraph{BLE}

@@

\paragraph{IMU}

@@

\paragraph{Servos}

@@

\paragraph{Protothreads}

@@

\paragraph{Gradient descent}

@@

\subsubsection{Testing}

\paragraph{BLE}

@@

\paragraph{IMU}

@@

\paragraph{Servos}

@@

\paragraph{Whole system}

@@

\subsection{Results}

@@

\subsection{Conclusions}

@@

\subsubsection{Appendix A: Data}

@@

\subsubsection{Appendix B: Bill Of Materials}

This project was done with a \$150 budget in mind. Costs for parts labeled ``Lab rental'' were borrowed from the department of Electrical and Computer Engineering at Cornell University, but can also be found at vendors like Digikey, Adafruit, and Amazon.
